Checkerboards And Missing Species Combinations: Are Ecological Communities

### 2016-02-07 ManuscriptMarkDown.Rmd
Updated figure references for final version.

So ecologists have often taken a different tack. They use so-called "natural experiments" in which they let nature do the work and generate the replicates. Islands in an archipelago are often treated as natural replicates. The patterns of species association among a set of islands may provide important clues for how species interact.

The data for such an analysis can be represented as a binary matrix in which each row is a species, each column is an island (or site), and the matrix entries represent the presence (1) or absence (0) of a particular species on a particular island. For a matrix M with i = 1 to S species and j = 1 to R replicate islands, M~ij~ indicates the presence or absence of species i on island j.

It is too much work to identify all of the different species of vertebrates, invertebrates, plant, and microbes on a set of islands, or even a single island. Instead, ecologists have usually restricted their analysis to a single taxonomic group of organisms that have similar body sizes and similar life histories. Such collections (or "ecological guilds") of species may often be competing for shared food or habitat resources.

@@

Connor and Simberloff argued that a randomization or reshuffling of the existing data matrix might mimic the properties of a null or non-interactive community, without getting bogged down in the problem of estimating colonization and extinction parameters.

Following standard Monte Carlo procedure, Connor and Simberloff randomized the observed presence-absence matrix, and calculated the number of checkerboard pairs and the number of species combinations that were found in this null matrix. Repeating this process 1000 times yields a distribution of the number of checkerboard pairs for this null hypothesis. Using a standard frequentist approach, we can then compare with the observed data and calculate how extreme or improbable the data are, given this null hypothesis as (*p*(Observed number of checkerboards | *H~0~*). The same procedure could be used to evaluate the number of species combinations that are found in the matrix.

@@

But exactly how should the observed data be reshuffled to mimic a colonization extinction model in which species occurrences on islands are independent of one another? The devil is in the details, and the development of different algorithms for randomizing a presence-absence matrix and quantifying the patterns has remained an active research frontier for over 35 years. The simplest initial approach is a kind of "isotropic" null hypothesis in which we simply reshuffle with equal probability all of the matrix elements. This null model preserves the dimensions of the matrix (the number of rows and columns) and it preserves, the "fill" of the matrix, that is the proportion of presences (0.17 for the West Indices finch matrix). But there are no other constraints on the occurrence of species on islands. Here is one such matrix, created by randomizing the occurrences in the original finch matrix:

@@

Here are the results for the 1000 randomizations of the West Indies finch matrix, with separate analyses for the number of checkerboard pairs and the number of species combinations. By this test, we would conclude that there are fewer checkerboard pairs than expected by chance, and also fewer observed combinations.

@@

Connor and Simberloff chose to impose both constraints, preserving simultaneously the row and column totals of the matrix to generate the expected number of checkerboards and observed species combinations. Figure 6 shows one of these simulated matrices, which is visually much more similar to the original matrix than random matrices produced by the other algorithms. Results for this *Fixed-Fixed* algorithm are qualitatively the same as for the *Fixed Row Sums* algorithm: fewer species combinations than expected by chance, but no evidence that the number of checkerboard species pairs is unusual. Notice also in the histogram scales that the range of simulated values is much less for the *Fixed-Fixed* algorithm than any of the other recipes, again reflecting the constraint of matching both the row and column sums to those of the original matrix.

@@

![The Markov chain swapping algorithm behaves like a sliding tile picture puzzle, in which consecutive random swaps to the binary presence absence matrix (Figure 1) obliterate the initial pattern.](Puzzle2.jpg)

@@

Randomization tests based only on the binary presence-absence matrix cannot tease apart these mechanisms. However, more recent approaches have begun to incorporate information on spatial pattern and habitat structure into the null models, and have even used an empirical Bayes approach to try and tease out the statistical significance of individual species pairs, as opposed to a single overall metric such as the number of checkerboard pairs for the entire matrix. In sum, the original controversy between Diamond and Connor and Simberloff over species co-occurrences on islands has yielded a set of intriguing statistical puzzles that continue to inspire ecologists as they interpret patterns of community assembly.

>>>>

So ecologists have often taken a different tack. They use so-called "natural experiments" in which they let nature do the work and generate the replicates. Islands in an archipelago, for example, are often treated as statistical replicates. The patterns of species association among a set of islands may provide important clues for how species interact.

The data for such an analysis can be represented as a binary matrix in which each row is a species, each column is an island (or site), and the matrix entries represent the presence (1) or absence (0) of a particular species on a particular island. For a matrix *M* with *i* = 1 to *S* species and *j* = 1 to *R* replicate islands, *M~ij~* indicates the presence or absence of species i on island *j*.

It is too much work to identify all of the different species of vertebrates, invertebrates, plant, and microbes on a single island, much less an entire archipelago. Instead, ecologists have usually restricted their analysis to a single taxonomic group of organisms that have similar body sizes and similar life histories. Such collections (or "ecological guilds") of species may often be competing for shared food or habitat resources.

@@

Connor and Simberloff argued that a randomization or reshuffling of the existing data matrix might mimic the properties of a null or non-interactive community, without getting bogged down in the problem of estimating colonization and extinction parameters.   Following standard Monte Carlo procedure, Connor and Simberloff randomized the observed presence-absence matrix, and calculated the number of checkerboard pairs and the number of species combinations that were found in this null matrix. Repeating this process 1000 times yields a distribution of the number of checkerboard pairs for this null hypothesis. Using a standard frequentist approach, Connor and Simberloff compared the observed and simulated data and calculated how extreme or improbable the data were, given this null hypothesis (*p*(Observed number of checkerboards | *H~0~*). The same procedure could be used to evaluate the number of species combinations --- or any other univariate metric ---  that are found in the matrix.

@@

But exactly how should the observed data be reshuffled to mimic a colonization extinction model in which species occurrences on islands are independent of one another? The devil is in the details, and the development of different algorithms for randomizing a presence-absence matrix and quantifying the patterns has remained an active research frontier for over 35 years. The simplest initial approach is a kind of "isotropic" null hypothesis in which we simply reshuffle with equal probability all of the matrix elements. This null model preserves the dimensions of the matrix (the number of rows and columns) and it preserves, the "fill" of the matrix, that is the proportion of presences (0.17 for the West Indices finch matrix). But there are no other constraints on the occurrence of species on islands. Figure 3 depicts one such matrix, created by randomizing the occurrences in the original finch matrix.

@@

Figure 4 depicts the results for the 1000 randomizations of the West Indies finch matrix, with separate analyses for the number of checkerboard pairs and the number of species combinations. By this test, we would conclude that there are fewer checkerboard pairs than expected by chance, and also fewer observed combinations.

@@

Connor and Simberloff chose to impose both constraints, preserving simultaneously the row and column totals of the original matrix to generate the expected number of checkerboards and observed species combinations. Figure 6 shows one of these simulated matrices, which is visually much more similar to the original matrix than random matrices produced by the other algorithms. Results for this *Fixed-Fixed* algorithm are qualitatively the same as for the *Fixed Row Sums* algorithm: fewer species combinations than expected by chance, but no evidence that the number of checkerboard species pairs is unusual. Notice also, in the histogram scales, that the range of simulated values is much less for the *Fixed-Fixed* algorithm than any of the other recipes, again reflecting the constraint of matching both the row and column sums to those of the original matrix.

@@

![The Markov chain swapping algorithm behaves like a sliding tile picture puzzle, in which consecutive random swaps to the binary presence absence matrix (Figure 1) obliterate the initial pattern.](Figure7_Gotelli.jpg)

@@

Randomization tests based only on the binary presence-absence matrix cannot tease apart these mechanisms. However, more recent approaches have begun to incorporate information on spatial pattern and habitat structure into the null models, and have even used an empirical Bayes approach to try and tease out the statistical significance of individual species pairs, as opposed to a single overall metric for the entire matrix such as the number of checkerboard pairs. In sum, the original controversy between Diamond and Connor and Simberloff over species co-occurrences on islands has yielded a set of intriguing statistical puzzles that continue to inspire ecologists as they interpret patterns of community assembly.
